Data Specialization for a Staged Lambda Calculus

### 2015-02-22 paper/intro.tex
finished a pass through the intro

\emph{Data specialization} \cite{knoblock96} is a different technique which
specializes $f$ \emph{without} the fixed argument $a$. Instead, it transforms
$f$ into a pair of functions $f_1$ and $f_2$. $f_1(a)$ produces a data structure
populated with the results of the subexpressions depending only on $a$. $f_2$
takes this data structure and the varying argument, and computes the result;
that is, $f_2(f_1(a),-)$ is the same as $f(a,-)$. Importantly, unlike in partial
evaluation, neither $f_1$ nor $f_2$ depend on $a$!

In this paper, we extend the technique of data specialization to a typed lambda
calculus with recursive and higher-order functions, and recursive datatypes.
Previous work has been limited to C-like languages (see graphics papers) and
doesn't handle certain stuff.

This additional expressive power allows us to perform data specialization on new
kinds of functions, like quickselect. We get some interesting results from
this---including algorithm synthesis with asymptotic speedup. Higher-order
functions also allow us to write combinators which cut across both stages (see
graphics stuff).

We start by walking through a quickselect example in \ref{sec:example}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm which performs data specialization.
Finally, in \ref{sec:examples}, we walk through what our algorithm does to
various programs, like deriving classic algorithms.

%It has potentially both control-flow and data dependencies on $a$.

%Many programs have arguments coming at different frequencies. It can be
%important to specialize these functions on some of their arguments, whether or
%not we know those arguments, in order to perform invariant computations only
%once.

%For example, quickselect. Or, the compilation of high-level shading languages
%for computer graphics such as RTSL~\cite{Proudfoot:2001} and
%Spark~\cite{Foley:2011}.

%\begin{itemize}
%\item Computing $b^e$, where $b$ varies and $e$ is constant.
%\item Finding the $i$th largest element in a list, where the list is constant
%and $i$ varies.
%\item Computing the shading of a surface in a 3D rendering, where some
%contributors to the final pixel color can be approximated as constant over
%pixel blocks.  
%\end{itemize}

>>>>

\emph{Data specialization} \cite{knoblock96} is a technique for specializing $f$
\emph{without} the fixed argument $a$, instead splitting $f$ into a pair of
functions $f_1$ and $f_2$. $f_1(a)$ produces a data structure containing the
results of the computations which depend only on $a$; $f_2$ then completes the
computation, given this data structure and the varying argument; that is,
$f_2(f_1(a),-)$ computes $f(a,-)$.

Previous work on data specialization has been limited to simple, imperative
languages. In this paper, we extend data specialization to a typed lambda
calculus, allowing us to specialize more interesting programs.

When splitting certain recursive functions, like \texttt{quickselect}, our
algorithm synthesizes recursive data structures and traversal algorithms which
yield asymptotic speedups over the original function. Splitting higher-order
combinators, like \texttt{map}, provides compositional reasoning at the source
level while cross-cutting fixed runtime stages, as in graphical shading
languages like Spark~\cite{Foley:2011}.

We start in \ref{sec:example} with an extended example of splitting
\texttt{quickselect}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm for performing data specialization.
Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
transforms a variety of other programs.
